# 536 R Recitation - 18 November

This lab session will go through essential data analysis techniques in R, including descriptive statistics, visualisation, data manipulation, and introductory inferential statistics. We will be using a dataset on **Song Sparrow cognition**.

## 1. Setup and Data Loading

First, we install and load the necessary R packages. The `tidyverse` is a collection of packages, including `dplyr` for data manipulation and `ggplot2` for plotting. We will also use `pwr` for power analysis and a custom function for the mode, as there is no built-in function in base R.

```{r}
#install.packages("BSDA")
# BSDA: Basic Statistics and Data Analysis
```

```{r}
# Install packages if you haven't already (uncomment the line below to run)
# install.packages(c("tidyverse", "pwr", "BSDA"))

# Load the required libraries
library(tidyverse)
library(readxl)
library(pwr)
library(BSDA) # For Z-test
library(DescTools)

# Load the dataset
df <- read_excel("song+sparrow+cognition+data.xlsx", sheet = "Sheet2")
```

### 1.1 Data Inspection: `str()` and `head()`

It's crucial to inspect the data structure and initial rows to understand the variables and check for any immediate issues.

#### `str()`

The `str()` (structure) function shows us the data type of each column and a sample of the data. This helps confirm if variables are being read as the correct type (e.g., numeric, character, or factor).

```{r}
# Inspect the structure of the data
str(df)
```

#### `head()`

The `head()` function displays the first six rows, which is a quick way to visually confirm the data has loaded correctly and to see the format of the entries.

```{r}
# View the first few rows of the data
head(df)
```

------------------------------------------------------------------------

## 2. Descriptive Statistics

Descriptive statistics summarize the main features of a dataset. We will focus on the **Neophobia** score (`neophobia`) which measures the latency (in seconds) for a bird to approach a novel object.

### 2.1 Selecting a Specific Column (Base R `$`)

In base R, you can select a specific variable (column) from a data frame using the dollar sign operator (`$`).

```{r}
# Select the 'neophobia' column
neophobia_scores <- df$neophobia
head(neophobia_scores)
```

### 2.2 Calculating Key Descriptives

We will calculate the requested descriptive statistics for the `neophobia` variable.

| Statistic | R Function | Explanation |
|:-----------------------|:-----------------------|:-----------------------|
| $N$ (Count) | `sum(!is.na())` or `length(na.omit())` | The number of non-missing observations. |
| **Mean** | `mean()` | The arithmetic average. |
| **Median** | `median()` | The middle value when the data is sorted. |
| **Mode** | `Mode()` | The most frequently occurring value. |
| **Range (Min/Max)** | `min()`, `max()`, `range()` | The smallest and largest values. |
| **Interquartile Range** | `IQR()` | The difference between the 75th and 25th percentiles ($Q3 - Q1$). |
| **Variance** | `var()` | The average of the squared differences from the Mean. |
| **Standard Deviation** | `sd()` | The square root of the variance, providing a measure of data spread in the original units. |

```{r}
# Descriptive Statistics for 'neophobia'

# N (Count)
n_obs <- length(na.omit(neophobia_scores))
cat("N (Count):", n_obs, "\n")

# Mean
mean_val <- mean(neophobia_scores, na.rm = TRUE)
cat("Mean:", round(mean_val, 2), "\n")

# Median
median_val <- median(neophobia_scores, na.rm = TRUE)
cat("Median:", median_val, "\n")

# Mode (using the custom function)
mode_val <- Mode(neophobia_scores)
cat("Mode(s):", mode_val, "\n")

# Min Value
min_val <- min(neophobia_scores, na.rm = TRUE)
cat("Minimum Value:", min_val, "\n")

# Maximum Value
max_val <- max(neophobia_scores, na.rm = TRUE)
cat("Maximum Value:", max_val, "\n")

# Range
range_val <- max_val - min_val
cat("Range (Max - Min):", round(range_val, 2), "\n")

# Interquartile Range (IQR)
iqr_val <- IQR(neophobia_scores, na.rm = TRUE)
cat("Interquartile Range (IQR):", round(iqr_val, 2), "\n")

# Variance
var_val <- var(neophobia_scores, na.rm = TRUE)
cat("Variance:", round(var_val, 2), "\n")

# Standard Deviation (SD)
sd_val <- sd(neophobia_scores, na.rm = TRUE)
cat("Standard Deviation (SD):", round(sd_val, 2), "\n")
```

------------------------------------------------------------------------

## 3. Base R Histogram

Histograms visualize the distribution of a numeric variable by dividing the data into bins and showing the frequency of observations in each bin.

### 3.1 Basic Histogram

```{r}
# Basic Histogram for 'neophobia'
hist(df$neophobia)
```

### 3.2 Modifying Titles and Bin Size

We can customize the histogram by changing the main title (`main`), axis labels (`xlab`, `ylab`), and the number or size of the bins (`breaks`).

```{r}
# Histogram with custom titles and bin size
hist(df$neophobia,
     main = "Distribution of Neophobia Scores in Song Sparrows", # Change Main Title
     xlab = "Neophobia Score (Latency in Seconds)",            # Change X-axis Label
     ylab = "Frequency of Birds (Count)",                      # Change Y-axis Label
     col = "lightblue",                                        # Change bar color
     border = "black",                                         # Add borders
     breaks = 10) # Set the number of bins to 10
```

------------------------------------------------------------------------

## 4. Probability Distributions (`dnorm`, `pnorm`, `qnorm`)

These functions are fundamental for working with the **Normal Distribution**. We'll use a hypothetical scenario: a population mean ($\mu$) of $100$ and a standard deviation ($\sigma$) of $15$.

### 4.1 `dnorm()`: Density

`dnorm(x, mean, sd)` calculates the **Probability Density** at a specific value $x$. This is the height of the curve at that point.

```{r}
# Calculate the density of a score of 115 (Z = 1)
dnorm(x = 115, mean = 100, sd = 15)
```

### 4.2 `pnorm()`: Cumulative Probability

`pnorm(q, mean, sd)` calculates the **Cumulative Probability** for a value $q$, which is the area under the curve to the left of $q$ (i.e., $P(X \le q)$).

```{r}
# Calculate the probability of a score being 115 or less
pnorm(q = 115, mean = 100, sd = 15)
```

### 4.3 `qnorm()`: Quantile (Inverse Cumulative Probability)

`qnorm(p, mean, sd)` calculates the **Quantile** (the $x$ value) that corresponds to a given cumulative probability $p$.

```{r}
# Calculate the score that corresponds to the 95th percentile (0.95 area to the left)
qnorm(p = 0.95, mean = 100, sd = 15)
```

------------------------------------------------------------------------

## 5. Data Type Conversion and Factor Level Reordering

Data types often need to be modified for correct analysis or representation. We use the `dplyr` function `mutate()` to create or change columns.

### 5.1 Converting to Numeric

We convert the `bird` variable (which is currently a `character` or `object` type) to a numeric type. *Note: This operation will result in `NA` values because the bird IDs contain letters (e.g., "sg701"), which is expected when converting non-numeric text to numeric.*

Examples that cause NA error :

-   `"sg701"`

-   `"bird_A"`

-   `"x12"`

-   `"blue"`

-   `"sg726"`

All of these contain **letters**, so `as.numeric()` cannot convert them to numbers.

Convert factor/character labels into numeric IDs: This produces:

| bird  | bird_id_numeric |
|-------|-----------------|
| sg701 | 1               |
| sg704 | 2               |
| sg707 | 3               |

No warning, because factor → numeric conversion is safe.

```{r}
df_converted <- df %>%
  mutate(bird_id_numeric = as.numeric(factor(bird)))

str(df_converted$bird_id_numeric)
head(df_converted$bird_id_numeric)
```

### 5.2 Converting from Character

Converting a variable like **sex** ("M"/"F") directly to numeric will typically result in missing values (\`NA\`) since the characters aren't numbers. However, by converting the variable to a **factor first**, R converts the underlying integer representation of the factor levels (e.g., \$1\$ for the first level, \$2\$ for the second) into a numeric column. This is an important concept in R data manipulation.

```{r}

# Convert 'sex' (character) -> Factor -> Numeric
# This reveals the underlying numerical representation of factor levels (e.g., 1, 2)
df_converted <- df %>%
  mutate(sex_numeric_level = as.numeric(factor(sex)))

# Inspect the structure (should now be 'dbl' or 'int')
str(df_converted$sex_numeric_level)

# View the first few values to see the level conversion
head(df_converted$sex_numeric_level)
```

### 5.3 Factor Level Reordering

**Factors** are used to store categorical data. We can change the order of levels using the `levels` argument in the `factor()` function. This is important for controlling the order in graphs (like bar plots) and statistical models.

Let's convert `sex` to a factor and then change the reference level from "M" to "F".

```{r}
# 1. Convert 'sex' to a factor
df <- df %>%
  mutate(sex_factor = factor(sex))

# See the default order of levels
levels(df$sex_factor)

# 2. Reorder the levels: place 'F' (Female) first, then 'M' (Male)
df <- df %>%
  mutate(sex_factor_reordered = factor(sex, levels = c("F", "M")))

# See the new order of levels
levels(df$sex_factor_reordered)
```

------------------------------------------------------------------------

## 6. `ggplot2` Visualizations

`ggplot2` is a powerful package for creating high-quality, customizable graphics. All plots follow the grammar of graphics: data $\rightarrow$ aesthetics (mapping variables to axes) $\rightarrow$ geometry (the plot type).

We will use the original `sex` column (M/F) and the `neophobia` score.

```{r}
# Ensure ggplot2 is loaded (it is part of tidyverse)
library(ggplot2)
```

### 6.1 Scatterplot

A scatterplot shows the relationship between two numeric variables. Let's plot `neophobia` against **Novel Foraging Time** (`novforag`).

```{r}
ggplot(df, aes(x = neophobia, y = novforag)) +
  geom_point(aes(color = sex)) + # Map 'sex' to color
  labs(
    title = "Neophobia vs. Novel Foraging Time",
    x = "Neophobia Score (s)",
    y = "Novel Foraging Time (s)",
    color = "Sex"
  ) +
  theme_minimal()
```

### 6.2 Bar Plot (`geom_bar` for Categorical Counts)

Used to visualize the count (frequency) of each category in a single categorical variable (`sex`).

```{r}
ggplot(df, aes(x = sex)) +
  geom_bar(fill = "darkorange") +
  labs(
    title = "Count of Birds by Sex",
    x = "Sex",
    y = "Count (n)"
  ) +
  theme_minimal()
```

### 6.3 Histogram (`geom_histogram`)

The `ggplot2` equivalent of the base R histogram.

```{r}
ggplot(df, aes(x = neophobia)) +
  geom_histogram(bins = 15, fill = "darkblue", color = "white") +
  labs(
    title = "Histogram of Neophobia Scores (ggplot2)",
    x = "Neophobia Score (s)",
    y = "Frequency"
  ) +
  theme_minimal()
```

### 6.4 Density Plot (`geom_density`)

A smoothed version of the histogram, which is excellent for visualizing the shape of the distribution.

```{r}
ggplot(df, aes(x = neophobia)) +
  geom_density(fill = "red", alpha = 0.5) + # alpha for transparency
  labs(
    title = "Density Plot of Neophobia Scores",
    x = "Neophobia Score (s)"
  ) +
  theme_minimal()
```

### 6.5 Boxplot (`geom_boxplot`)

Displays the distribution of a numeric variable (e.g., `neophobia`) across different categories (e.g., `sex`). It shows the median, quartiles, and outliers.

```{r}
ggplot(df, aes(x = sex, y = neophobia)) +
  geom_boxplot(fill = "green", alpha = 0.7) +
  labs(
    title = "Boxplot of Neophobia Scores by Sex",
    x = "Sex",
    y = "Neophobia Score (s)"
  ) +
  theme_minimal()
```

### 6.6 Violin Plot (`geom_violin`)

A violin plot is similar to a boxplot but also shows the kernel density of the data at different values (providing a better view of the distribution shape).

```{r}
ggplot(df, aes(x = sex, y = neophobia)) +
  geom_violin(aes(fill = sex), alpha = 0.5) +
  geom_boxplot(width = 0.1) + # Add a small boxplot inside
  labs(
    title = "Violin Plot of Neophobia Scores by Sex",
    x = "Sex",
    y = "Neophobia Score (s)"
  ) +
  theme_minimal()
```

------------------------------------------------------------------------

## 7. Data Filtering, Grouping, and Summarisation (`dplyr`)

The `dplyr` package is the core tool for data manipulation in the `tidyverse`.

### 7.1 Data Filtering (`filter()`)

The `filter()` function selects a subset of rows based on specified logical conditions.

```{r}
# Filter the data to include only Male birds ('M')
df_males <- df %>%
  filter(sex == "M")

# Check the new dataset size and head
cat("Number of Male birds:", nrow(df_males), "\n")
head(df_males)
```

### 7.2 Data Grouping and Summarisation (`group_by()` and `summarise()`)

We can group the data by a categorical variable (`sex`) and then calculate summary statistics for a numeric variable (`neophobia`) within each group.

```{r}
# Group by 'sex' and calculate descriptive statistics for 'neophobia'
df_summary <- df %>%
  group_by(sex) %>%
  summarise(
    N = n(),
    Mean_Neophobia = mean(neophobia, na.rm = TRUE),
    Median_Neophobia = median(neophobia, na.rm = TRUE),
    SD_Neophobia = sd(neophobia, na.rm = TRUE),
    Min_Neophobia = min(neophobia, na.rm = TRUE),
    Max_Neophobia = max(neophobia, na.rm = TRUE)
  )

# View the summary table
print(df_summary)
```

------------------------------------------------------------------------

## 8. Missing Data (`NA`) Handling

Missing values (`NA`) can affect analysis. We must identify and decide how to treat them. We will use the `colassoc` column, which has 2 missing values.

### 8.1 Counting `NA` Values

We use `is.na()` to check for missing values and `sum()` to count them.

```{r}
# Count NA values in the whole dataset
cat("Total NA values in the dataset:", sum(is.na(df)), "\n")

# Count NA values in a specific column ('colassoc')
na_count_colassoc <- sum(is.na(df$colassoc))
cat("NA count in 'colassoc':", na_count_colassoc, "\n")
```

### 8.2 Filtering/Removing Rows with `NA` (`drop_na()`)

We can remove rows that contain `NA` values in specific columns using `drop_na()` from `tidyr` (part of `tidyverse`).

```{r}
# Remove rows with NA in 'colassoc'
df_clean_colassoc <- df %>%
  drop_na(colassoc)

# Check the new size of the dataset
cat("Original rows:", nrow(df), "\n")
cat("Rows after removing NA in 'colassoc':", nrow(df_clean_colassoc), "\n")

# Remove rows with NA in *any* column
df_complete_cases <- df %>%
  drop_na()
  
cat("Rows after removing NA in *any* column:", nrow(df_complete_cases), "\n")
```

------------------------------------------------------------------------

## 9. Hypothesis Testing and Normality

We conclude with fundamental inferential statistics and a check for data normality. We'll use the `neophobia` score.

### 9.1 One-Sample Z-Test

A Z-test is used when the population standard deviation ($\sigma$) is known. We'll assume the population $\sigma$ for neophobia is $50.0$ and we test if the mean is significantly different from a hypothesized population mean ($\mu_0$) of $60$ seconds.

```{r}
# One-Sample Z-Test
# Null Hypothesis (H0): Population mean neophobia is 60 seconds (mu = 60)
# Alternative Hypothesis (HA): Population mean neophobia is NOT 60 seconds (mu != 60)

# We use the z.test function from the BSDA package
z.test(
  x = df$neophobia,
  mu = 60,       # Hypothesized population mean
  sigma.x = 50,  # Known population standard deviation (assumed)
  conf.level = 0.95
)
```

### 9.2 One-Sample T-Test

A T-test is used when the population standard deviation is **unknown** (which is more common). We test if the mean neophobia score is significantly different from a hypothesized value ($\mu_0$) of $50$ seconds.

```{r}
# One-Sample T-Test
# Null Hypothesis (H0): Population mean neophobia is 50 seconds (mu = 50)
# Alternative Hypothesis (HA): Population mean neophobia is NOT 50 seconds (mu != 50)

t.test(
  x = df$neophobia,
  mu = 50,       # Hypothesized population mean
  conf.level = 0.95)
```

### 9.3 Sample Size Calculation for One-Sample T-Test

The `pwr` package can determine the required sample size ($N$) to detect a certain effect size ($\delta$) with a specified power ($1 - \beta$) and significance level ($\alpha$).

We want to find $N$ to detect a **medium** effect size ($\delta = 0.5$) with $80\%$ power ($\beta = 0.20$) at an $\alpha = 0.05$ significance level.

```{r}
# Calculate required sample size (n) for a one-sample t-test
pwr.t.test(
  d = 0.5,        # Effect size (Cohen's d: 0.2=small, 0.5=medium, 0.8=large)
  sig.level = 0.05, # Significance level (alpha)
  power = 0.80,   # Desired power (1 - beta)
  type = "one.sample",
  alternative = "two.sided")
```

### 9.4 QQ Plot (Quantile-Quantile Plot)

A QQ plot is a visual tool to assess if a dataset is approximately normally distributed. If the data is normal, the points will fall along the straight diagonal line.

```{r}
# Create a QQ plot for 'neophobia'
qqnorm(df$neophobia,
       main = "QQ Plot of Neophobia Scores",
       xlab = "Theoretical Quantiles (Normal Distribution)",
       ylab = "Sample Quantiles (Neophobia Score)")
qqline(df$neophobia, col = "red") # Add a reference line
```

### 9.5 Shapiro-Wilk Test (Normality Testing)

The Shapiro-Wilk test is a formal statistical test for normality.

-   **Null Hypothesis (**$H_0$): The data is drawn from a normally distributed population.
-   **Alternative Hypothesis (**$H_A$): The data is **not** drawn from a normally distributed population.

If the $p$-value is less than the significance level (e.g., $\alpha = 0.05$), we reject $H_0$ and conclude the data is **not** normally distributed.

```{r}
# Perform the Shapiro-Wilk test on 'neophobia' scores
shapiro.test(df$neophobia)
```

------------------------------------------------------------------------

*Note: For all hypothesis tests, interpret the results based on the p-value. If* $p < \alpha$, the result is statistically significant.
